---
title: "Class 3 Cox Model"
author: "Yubo Rasmussen"
format: html
editor: visual
---

# Introduction

This notebook demonstrates how to fit and analyze a Cox proportional hazards model in R using the `survival` package. Additionally, it explores numerical derivatives, partial log-likelihood, and hypothesis tests such as the Wald, Likelihood Ratio Test (LRT), and Score tests. Along the way, questions are posed to deepen understanding of the methods and their applications.

## Example: Cox Model Estimate with `coxph()`

### Load and Prepare the Data

```{r}
library(survival)

# Time to event (in arbitrary units)
Time <- c(2,5,9,11,14, 4,7,10,12,15)

# Censoring indicator (1 = event occurred, 0 = censored)
Censor <- c(1,0,1,0,1, 1,0,0,0,1)

# Grouping variable (1 = Male, 2 = Female)
Sex <- c(rep(1,5), rep(2,5))

# Display the dataset
cbind(Time, Censor, Sex)

# Create a survival object
Surv(Time, Censor)
```

**Questions and Answers**:

1\. **What does the `Surv()` function do, and why is it essential in survival analysis?**

2\. **Why do we include a censoring indicator, and how does it affect the analysis?**

### Fit the Cox Model

```{r}
# Fit the Cox model
Cox.fit <- coxph(Surv(Time, Censor) ~ Sex)

# Display the model summary
summary(Cox.fit)
```

**Questions and Answers**:

1\. **What does the coefficient for `Sex` represent in the Cox model?**

2\. **How would you interpret the hazard ratio derived from the coefficient?** -

3\. **What does the p-value in the output tell us about the relationship between `Sex` and survival time?**

------------------------------------------------------------------------

## Partial Log-Likelihood and Score Function

### Define the Partial Log-Likelihood

```{r}
PartialPL <- function(b){
  p <- exp(b)
  b - 3*log(1 + p) - log(4 + 5*p)
}
```

**Questions and Answers**:

1\. **What does the partial log-likelihood represent in the context of the Cox model?**

2\. **How does the log-likelihood depend on the parameter `b` (beta)?**

### Plot the Partial Log-Likelihood

```{r}
PlotPL <- function(Left, Right){
  Beta <- seq(Left, Right, length = 1000)
  plot(Beta, PartialPL(Beta), type = "l", main = "Partial Log Likelihood",
       xlab = "Beta", ylab = "Log Likelihood", col = "red", lwd = 2)
  grid()
}

# Example plots
PlotPL(-2, 2)
PlotPL(-1.157, -1.156)

# Estimated beta
Beta.hat <- -1.16
```

**Questions and Answers**:

1\. **What does the maximum of the log-likelihood plot indicate?**

2\. **How would narrowing the range of `Beta` values improve the accuracy of locating the maximum?**

### Define and Plot the Score Function

```{r}
Score <- function(b){
  p <- exp(b)
  1 - 3*p/(1 + p) - 5*p/(4 + 5*p)
}

PlotScore <- function(Left, Right){
  Beta <- seq(Left, Right, length = 1000)
  plot(Beta, Score(Beta), type = "l", main = "Score Function",
       xlab = "Beta", ylab = "Score", col = "red", lwd = 2)
  grid()
}

# Plot the score function
PlotScore(-1.157, -1.156)
```

**Questions and Answers**:

1\. **How is the score function related to the log-likelihood function?**

### Numerical Derivatives and Standard Error

```{r}
# Numerical derivative at Beta.hat
Derive <- (Score(Beta.hat + 0.01) - Score(Beta.hat - 0.01)) / 0.02

# Standard error
SE <- sqrt(-1 / Derive)
SE
```

------------------------------------------------------------------------

## Numerical Derivatives and Tests

### Numerical Derivatives

```{r}
PartialL <- function(x){
  y <- exp(x)
  x - 3*log(1 + y) - log(4 + 5*y)
}

Deriv <- function(x, h) (PartialL(x + h/2) - PartialL(x - h/2)) / h
Deriv2 <- function(x, h) (PartialL(x + h) - 2*PartialL(x) + PartialL(x - h)) / h^2

# Calculate derivative and second derivative
Info <- -Deriv2(Beta.hat, 10^(-4))
Var <- 1 / Info
Info
```

**Questions and Answers**:

1\. **How do numerical derivatives approximate the gradient and curvature of the log-likelihood function?**

2\. **What are the implications of the information matrix for parameter estimation?**

### Wald, LRT, and Score Tests

```{r}
# Wald test
W <- (Beta.hat - 0)^2 / Var

# Likelihood Ratio Test (LRT)
Lambda <- -2 * (PartialL(0) - PartialL(Beta.hat))

# Score test
U0 <- Deriv(0, 10^(-4))
I0 <- -Deriv2(0, 10^(-4))
Score <- U0^2 / I0

# Combine results
Tests <- c(W, Lambda, Score)
names(Tests) <- c("Wald", "LRT", "Score")
Tests
```

**Questions and Answers**:

1\. **What is the null hypothesis tested by each of the Wald, LRT, and Score tests?**

2\. **How are these three tests related, and why might they yield similar or different results?** -

3\. **What does the test statistic from each test imply about the significance of `Beta.hat`?**
