---
title: "Class 5 GLMs and Graduation"
author: Yubo Rasmussen and Alistair Wallis
format: docx
editor: visual
---

## Introduction

This document explores Gompertz's law of mortality using different modeling approaches, including normal models, Poisson regression, and binomial models (GLMs), applied to CMI data.

Let us first load the data, ensuring that the files CMI_Deaths.csv and CMI_Exposures.csv are inside your working directory. You can use the R script CMI_read.r on Canvas which will read in the datasets and tidy them up.

```{r}
source("CMI_read.r")
```

## Normal Models

We begin by fitting a normal model, i.e. a linear regression model, to the logarithm of observed forces of mortality, after first assuming our observations of deaths for each age come from a Poisson model of mortality. Can you recall what this looks like mathematically?

### Select Data for Year 2000, Ages 40-80

Your Task first is to select and store the number of deaths and exposure for ages 40-80 in 2000. Define a new variable "Obs" that adjusts for exposure and plot "Age" against "Obs".

```{r}
DTH <- Dth[ (40 <= Age) & (Age <= 80), Year == 2000] 
EXP <- Exp[ (40 <= Age) & (Age <= 80), Year == 2000] 
AGE <- 40:80 
Obs <- log(DTH/EXP) 
plot(AGE, Obs, xlab = "Age", ylab = "log(mortality)", main = "Gompertz Law: CMI Data, Ages 40-80, Year 2000") 
```

Q: The Gompertz law of mortality states that mortality rates increase exponentially with age. Why do we use `Obs <- log(DTH/EXP)`?

### Fit Simple Normal Model

Let us proceed to fit a simple normal model:

$$
\log(\text{Obs}) = \beta_0 + \beta_1\text{Age} + \varepsilon_i
$$

Using `lm`, proceed to fit `Obs ~ AGE`. Add the line of best fit from your regression model to the plot above.

```{r}
Gomp.fit <- lm(Obs ~ AGE) 
# Alternatively as a GLM: Gomp.fit <- glm(Obs ~ AGE, family = gaussian)
names(Gomp.fit) 
```

```{r}
plot(AGE, Obs, xlab = "Age", ylab = "log(mortality)", 
     main = "Gompertz Law: CMI Data, Ages 40-80, Year 2000") 
lines(AGE, Gomp.fit$fit, lwd = 2)
```

What are the coefficients of the fitted model?

```{r}
Gomp.fit$coefficients
```

How would you interpret this model and what is $\varepsilon_i$?

### Fit Weighted Normal Model

A simple normal model assumes that all observations are equally important, whereas a weighted normal model says otherwise. The weighted model assigns weights to observations based on the number of deaths, where ages with more deaths have a higher effect on the regression line, and vice versa.

Now you will proceed to fit a weighted normal model, and plot it as before. Are there any differences in the results between the two models?

```{r}
Gomp.fit.w <- lm(Obs ~ AGE, weights = DTH) 
 
plot(AGE, Obs, xlab = "Age", ylab = "log(mortality)", 
     main = "Gompertz Law: CMI Data, Ages 40-80, Year 2000") 
lines(AGE, Gomp.fit.w$fit, lwd = 2, lty = 2) 
text(42, -3.5, "log(m) = -11.81 + 0.111 * Age, weighted", adj = 0, cex = 1.4) 
```

### Residual Analysis

Residuals represent the difference between the actual observed values and the predicted values from our model. Examining residuals helps us understand the goodness of fit and whether the assumptions of the model hold.

1.  **Compute the standard deviation of residuals**: This is used to standardize the residuals.
2.  **Compute the standardised residuals**: These are defined as $z_i = \frac{y_i - \hat{y}_i}{\sigma}$, where $\sigma$ is the residual standard deviation. \[Think about how this would be adjusted for the **weighted** linear model\]
3.  **Plot residuals**: This helps identify potential outliers and checks if residuals are randomly distributed (which supports model validity).

First, let us consider the model without weights:

```{r}
Sigma <- sqrt(sum(Gomp.fit$res^2)/Gomp.fit$df.res) 
# Alternatively: Sigma <- summary(Gomp.fit)$sigma
Z.vals <- (Obs - Gomp.fit$fit)/Sigma 
# Alternatively: 
# Z.vals <- Gomp.fit$res/Sigma
# Z.vals <- resid(Gomp.fit, type = "pear")/Sigma
# Z.vals <- resid(Gomp.fit, type = "deviance")/Sigma
plot(AGE, Z.vals, ylab = "Standardized Residuals", 
     main = "Residuals in Gompertz Model: Normal Errors") 
abline(h = 0, lty = 2, col = "red") 
```

**What are your Interpretations of the residuals?**

-   If the residuals are randomly scattered around zero, the model’s assumptions are likely valid, it doesn’t look like the normal assumption is valid here. 
-   Systematic patterns in the residual plot suggest potential model misspecifications, it seems like there is a quadratic pattern. 
-   Large residuals may indicate outliers or data issues, which is not an issue here. 

**More advanced topics:**

For those also studying F79MB Statistical Models B, try to calculate the following:

**Residual deviance:**

```{r}
sum((Obs - Gomp.fit$fit)^2)
# Alternatively:
# Gomp.fit_GLM <- glm(Obs ~ AGE, family = gaussian)
# R = resid(Gomp.fit_GLM, type = "deviance")
# sum(R^2)
# or simply:
# Gomp.fit_GLM$deviance
```

**Null deviance:**

```{r}
ave_y = mean(Obs)
sum_of_squares = sum((Obs - ave_y)^2)
sum_of_squares
# check: summary(Gomp.fit_GLM)$null
```

**AIC:**

```{r}
n = length(Obs)
AIC_LM = sum((Obs - Gomp.fit$fit)^2)/Sigma^2 + 2*n*log(Sigma*sqrt(2*pi)) + 6
AIC_LM
# Alternatively: summary(Gomp.fit_GLM)$aic
# Note slight discrepancy
```

## Poisson Model

In this section, we model mortality rates using a **Poisson Generalised Linear Model (GLM)** with a **log-link function**. The Poisson model assumes that the number of deaths follows a Poisson distribution:

$$
D_x \sim \text{Poisson}(\lambda_x)
$$

where the **logarithm of the expected deaths** is modeled as a **linear function of age**:

$$
\log(\lambda_x) = \beta_0 + \beta_1 x + \log(\text{EXP}_x)
$$

Here, $\text{EXP}_x$ is the exposure (i.e., number of life-years at risk), included as an **offset** in the model.

### Fit the Poisson GLM Model

```{r}
Gomp.glm <- glm(DTH ~ offset(log(EXP)) + AGE, family = poisson) 
summary(Gomp.glm) 
```

-   The `offset(log(EXP))` ensures that the model adjusts for exposure.
-   The estimated coefficients $\beta_0$ and $\beta_1$ determine the mortality curve.

### Interpret the Coefficients

-   The **intercept (**$\beta_0$) represents the log mortality rate at the baseline age.
-   The **slope (**$\beta_1$) determines how mortality increases with age.
-   A positive slope indicates that mortality increases exponentially over time, consistent with **Gompertz’s Law**.

------------------------------------------------------------------------

### Plot the results for visualisation

Visualise the **observed mortality rates** and compare them with **model predictions**.

```{r}
plot(AGE, Obs, xlab = "Age", ylab = "log(mortality)", 
     main = "Gompertz Law: Poisson Model, CMI Data, Ages 40-80") 
 
# Add the fitted Poisson regression line 
lines(AGE, Gomp.glm$coef[1] + Gomp.glm$coef[2]*AGE, col = "green", lwd = 2) 
lines(AGE, log(Gomp.glm$fit/EXP), col = "blue", lwd = 2) 
 
# Add a legend for clarity 
legend("topleft", legend = c("Poisson Model (Linear Predictor)",  
                             "Poisson Model (Fitted Values)"),  
       col = c("green", "blue"), lty = c(1, 1), lwd = 2)
```

#### Alternative Visualisation: Comparing Poisson and Normal Model

Since we already fitted a **normal model (OLS regression)**, it is useful to **compare** it with the Poisson model.

```{r}
plot(AGE, Obs, xlab = "Age", ylab = "log(mortality)", 
     main = "Comparison: Normal vs Poisson Model") 
 
lines(AGE, Gomp.fit$fit, col = "red", lwd = 2, lty = 2)  # Normal Model 
lines(AGE, log(Gomp.glm$fit/EXP), col = "blue", lwd = 2)  # Poisson Model 
 
legend("topleft", legend = c("Normal Model", "Poisson Model"), 
       col = c("red", "blue"), lty = c(2,1), lwd = 2)  
```

------------------------------------------------------------------------

### Residual Analysis for Poisson Model

Residuals help us **assess model fit** and detect possible **misspecifications**.

#### Plot the Pearson and Deviance Residuals

-   **Pearson residuals** measure how far observed values deviate from expected values, standardized by variance.
-   **Deviance residuals** quantify goodness of fit by comparing observed and fitted values.

```{r}
plot(AGE, resid(Gomp.glm, type = "pear"), xlab = "Age", 
     ylab = "Residuals", main = "Residuals in Gompertz Model: Poisson") 
points(AGE, resid(Gomp.glm, type = "deviance"), pch = 2, col = "red") 
legend(50, 4, legend = c("Pearson Residuals", "Deviance Residuals"),  
       pch = c(1, 2), col = c("black", "red")) 
```

**More advanced**

Try to calculate the Pearson and deviance residuals yourself

```{r}
# Pearson
(DTH - Gomp.glm$fit)/sqrt(Gomp.glm$fit)
```

```{r}
# deviance residuals
2*(DTH*(log(DTH)-Gomp.glm$lin)-DTH+Gomp.glm$fit)
```

#### Alternative Residual Analysis: Q-Q Plot

A **Q-Q plot** compares the residuals to a normal distribution to check for **deviation from normality**.

```{r}
qqnorm(resid(Gomp.glm, type = "pear"), main = "Q-Q Plot: Poisson Residuals") 
qqline(resid(Gomp.glm, type = "pear"), col = "red") 
```

------------------------------------------------------------------------

#### What are potential model refinements?

If residuals indicate poor fit: 

1\. **Quadratic Model:** Add an $AGE^2$ term to allow for curvature. 

2\. **Binomial Model:** If overdispersion is present, a **binomial** GLM may be more appropriate. 

3\. **Spline Regression:** A **generalised additive model (GAM)** could capture nonlinear trends. 

**More advanced topics:**

For those also studying F79MB Statistical Models B, try to calculate the following:

**Residual deviance:**

```{r}
Dev_Poi <- 2*(DTH*(log(DTH)-Gomp.glm$lin)-DTH+Gomp.glm$fit)
sum(Dev_Poi)
# Alternatively:
# check: Gomp.glm$deviance
```

**Null deviance:**

```{r}
K = sum(DTH)/sum(EXP)
null_dev_poi <- sum(2*(DTH*(log(DTH)-log(EXP*K))-DTH+EXP*K))
null_dev_poi
# Alternatively, fit the null model:
# null_glm <-glm(DTH ~ offset(log(EXP)), family = poisson)
# null_dev_poi <- sum(2*(DTH*(log(DTH)-log(null_glm$fit))-DTH+null_glm$fit))
# check: summary(Gomp.glm)$null
```

**AIC:**

```{r}
log_fact_pois = rep(0,41)
for (j in 1:41){
  for (i in 1:DTH[j]){
    log_fact_pois[j] = log_fact_pois[j]+log(i)
  }
}
AIC_pois <- 4 - 2*sum(DTH*Gomp.glm$lin - Gomp.glm$fit - log_fact_pois)
AIC_pois
# Check: summary(Gomp.glm)$aic
```

## 

## Binomial Model

Now, we move to the **Binomial Generalised Linear Model (GLM)** using a **logit-link function**. This model assumes that the number of deaths follows a **binomial distribution**:

$$
D_x \sim \text{Binomial}(E_x, q_x)
$$

and we are aiming to model

$$
\tilde{q}_{x} = \frac{D_{x}}{E_{x}}
$$

where the **logit transformation** is applied to mortality probabilities:

$$
\log\left(\frac{q_x}{1 - q_x}\right) = \beta_0 + \beta_1 x
$$

### Fit the Binomial Model

```{r}
E.init <- EXP + DTH/2  # Initial exposure to risk 
Q.x <- DTH / E.init  # Mortality probability estimate 
Logit <- function(x) log(x / (1 - x))  # Define logit function 
Obs <- Logit(Q.x)  # Transform mortality rates 
 
# Fit the Binomial GLM 
Gomp.bin <- glm(Q.x ~ AGE, weights = E.init, family = binomial) 
summary(Gomp.bin) 
```

### Visualise the Binomial Model

```{r}
plot(AGE, Obs, xlab = "Age", ylab = "logit(mortality)", 
     main = "Gompertz Law: Binomial Model, Logit Link") 
 
# Add the fitted Binomial regression line 
lines(AGE, Gomp.bin$coef[1] + Gomp.bin$coef[2] * AGE, col = "red", lwd = 2) 
lines(AGE, Logit(Gomp.bin$fit), col = "blue", lwd = 2) 
 
legend("topleft", legend = c("Linear Predictor", "Fitted Values"), 
       col = c("red", "blue"), lty = c(1, 1), lwd = 2) 
```

### Perform Residual Analysis for Binomial Model

```{r}
plot(AGE, resid(Gomp.bin, type = "pear"), xlab = "Age", 
     ylab = "Residuals", main = "Residuals in Gompertz Model: Binomial") 
points(AGE, resid(Gomp.bin, type = "deviance"), pch = 2, col = "red") 
legend(50, 4, legend = c("Pearson Residuals", "Deviance Residuals"),  
       pch = c(1, 2), col = c("black", "red")) 
```

### Interpret the Residuals

-   If residuals are **randomly distributed**, the model is a good fit. 
-   **Patterns in residuals** suggest a need for a more flexible model (e.g., a quadratic term or splines). 
-   Large residuals may indicate **outliers or data misfits**. 

**More advanced topics:**

Try to calculate the Pearson and deviance residuals yourself:

```{r}
# Pearson
(DTH - E.init*Gomp.bin$fit)/sqrt(E.init*Gomp.bin$fit)
```

```{r}
# deviance residuals
2*(DTH*(Logit(Q.x)-Gomp.bin$lin)+E.init*(log(1-Q.x)-log(1-Gomp.bin$fit)))
```

For those also studying F79MB Statistical Models B, try to calculate the following:

**Residual deviance:**

```{r}
Dev_bin <- 2*(DTH*(Logit(Q.x)-Gomp.bin$lin)+E.init*(log(1-Q.x)-log(1-Gomp.bin$fit)))
sum(Dev_bin)
# Alternatively:
# check: Gomp.bin$deviance
```

**Null deviance:**

```{r}
qx_full <- sum(DTH)/sum(E.init)
null_dev_bin <- sum(2*(DTH*(Logit(Q.x)-Logit(qx_full))+E.init*(log(1-Q.x)-log(1-qx_full))))
null_dev_bin
# check: summary(Gomp.bin)$null
```

**AIC:**

```{r}
log_prod_bin = rep(0,41)
for (j in 1:41){
  for (i in 1:DTH[j]){
    log_prod_bin[j] = log_prod_bin[j]+log((E.init[j] - i + 1)/i)
  }
}
AIC_bin <- 4 - 2 * sum(DTH*Gomp.bin$lin + E.init*log(1-Gomp.bin$fit)+log_prod_bin)
AIC_bin
# Check: summary(Gomp.bin)$aic
```

## 

------------------------------------------------------------------------

## Gompertz Models Over Time

We now extend the Gompertz model to analyze how mortality changes over time for a given age.

```{r}
DTH <- Dth[Age == 65, ] 
EXP <- Exp[Age == 65, ] 
Obs <- log(DTH / EXP) 
 
plot(Year, Obs, xlab = "Year", ylab = "log(mortality)", 
     main = "Gompertz Style Models Over Time: Age 65") 
 
Gomp.yr <- glm(DTH ~ offset(log(EXP)) + Year, family = poisson) 
lines(Year, log(Gomp.yr$fit / EXP), lwd = 2, col = "blue") 
```

### Quadratic Model for Time Trends

```{r}
Year2 <- (Year - mean(Year))^2 
Gomp.yr.2 <- glm(DTH ~ offset(log(EXP)) + Year + Year2, family = poisson) 
plot(Year, Obs, xlab = "Year", ylab = "log(mortality)", 
     main = "Gompertz Style Models Over Time: Age 65") 
lines(Year, log(Gomp.yr$fit / EXP), lwd = 2, col = "blue") 
 
lines(Year, log(Gomp.yr.2$fit / EXP), lwd = 2, lty = 2, col = "red") 
legend("bottomleft", legend = c("Linear Model", "Quadratic Model"), 
       col = c("blue", "red"), lty = c(1, 2), lwd = 2) 
```

------------------------------------------------------------------------

### Residual Analysis for Time Trends

```{r}
plot(Year, resid(Gomp.yr, type = "pear"), xlab = "Year", 
     ylab = "Residuals", main = "Residuals for Time-Based Model") 
points(Year, resid(Gomp.yr.2, type = "pear"), pch = 2, col = "red") 
legend("topright", legend = c("Linear Model Residuals", "Quadratic Model Residuals"), 
       pch = c(1, 2), col = c("black", "red")) 
```

------------------------------------------------------------------------

## **Conclusion**

-   The **Poisson model** works well but may suffer from overdispersion.
-   The **Binomial model** provides an alternative framework based on survival probabilities.
-   Time-based models suggest that mortality rates change over time and may require **nonlinear adjustments**.
