---
title: "Class 4 (Solution)"
author: "Yubo Rasmussen"
format: html
editor: visual
---

## Introduction

This notebook covers the Binomial Model with the Uniform Distribution of Deaths (UDD) and Constant Force of Mortality (CFM) assumptions. Additionally, it implements the Newton-Raphson method for numerical optimization.

------------------------------------------------------------------------

## Binomial Model: Uniform Distribution of Deaths (UDD)

Write down the loglikelihood function from Example 4.7 in terms of q and assign it as a named variable under the UDD assumption.

```{r}
# Log likelihood function in terms of q

```

Write a function to plot the log-likelihood function above and call it.

```{r}
# Plot log likelihood

# Plotting
```

What are we looking for in this plot?

What do you think is a reasonable assumption for the value of `Qhat`? Assign and store it.

```{r}
```

Write down the score function of Q. Can you explain how the score function is derived?

```{r}
# Score for Q
```

Now write a function to plot the Score function and call it.

```{r}
# Plot of Score function
```

What are we looking for in this plot?

What do you think is a reasonable assumption for the value of `QHat`?

```{r}
```

Now, it is time to compute the standard error of the estimated parameter `QHat`.

Use a numerical approximation by computing the difference quotient.

```{r}
# SE numerical
```

Use an analytical second derivative approach.

```{r}
#   And with second derivative
```

Do the two estimates of SE agree?

------------------------------------------------------------------------

## Binomial Model: Constant Force of Mortality (CFM)

Write down the loglikelihood function from Example 4.7 in terms of $\mu$ and assign it as a named variable under the CFM assumption.

```{r}
# Log likelihood function in terms of mu
```

Write a function to plot the log-likelihood function above and call it.

```{r}
# Plot log likelihood

# Plotting

```

What are we looking for in this plot?

What do you think is a reasonable assumption for the value of `MuHat`? Assign and store it.

```{r}
```

Write down the loglikelihood function from Example 4.7 in terms of $\phi = \exp(-\mu/4)$ and assign it as a named variable under the CFM assumption.

```{r}
# Log likelihood function in terms of phi
```

Write a function to plot the log-likelihood function above and call it.

```{r}
# Plot log likelihood for Phi
```

What are we looking for in this plot?

What do you think is a reasonable assumption for the value of `PhiHat`? Assign and store it.

```{r}
```

Now check the consistency, i.e. is $\phi = \exp(-\mu/4)$ close to $\hat{\phi}$

```{r}

```

Write down the score function for Phi. Can you explain how the score function is derived for Phi?

```{r}
# Score for Phi
```

Now write a function to plot the Score function and call it.

```{r}
# Plot of Score function
```

What are we looking for in this plot?

What do you think is a reasonable assumption for the value of `PhiHat`?

```{r}
```

Now, it is time to compute the standard error of the estimated parameter `PhiHat`.

Use a numerical approximation by computing the difference quotient.

```{r}
# SE numerical
```

Use an analytical second derivative approach.

```{r}
#   And with second derivative
```

Do the two estimates of SE agree?

Repeat the above and write down the Score function and SE for mu, including a plot of the Score function.

```{r}
# Score function and SE for mu
```

What can you say about the relationship between mu and phi? Why do we compute both of them? Which score function do you think is easier to derive analytically?

------------------------------------------------------------------------

## Newton-Raphson Method

Newton-Raphson is an iterative method used to solve equations of the form $f(x) = 0$. The update rule for finding a better estimate of the root is given by:

$x_{n+1} = x_n - \frac{f(x_n)}{f'(x_n)}$

Since computing the analytical derivative $f'(x)$ is not always convenient, we use a numerical approximation.

### Implementing the Newton-Raphson Method

```{r}
# Newton-Raphson iteration using numerical differentiation
# Arguments: f, a function defining the equation
#           x0, an estimate of the solution
#           Delta, h-value for numerical differentiation
Newton <- function(f, x0, Delta){ 
  Derivative <- (f(x0+Delta/2) - f(x0-Delta/2))/Delta 
  x0 - f(x0)/Derivative 
} 
```

### Example: Finding the Square Root of 2

```{r}
Square <- function(x) x^2 - 2 
x1 <- Newton(Square, 1.5, 0.0001) 
x2 <- Newton(Square, x1, 0.0001) 
c(x1, x2, x2^2) 
```

### Generalizing Newton-Raphson with a Loop

We can now define a function that iteratively applies the Newton-Raphson method until convergence:

```{r}
# Additional argument: Tol, tolerance for solution
#                      Max.iter, maximum number of iterations
Solve <- function(f, x0, Delta, Tol, Max.Iter){ 
  TOL <- abs(f(x0)) 
  Iter <- 0 
  while((TOL > Tol) && (Iter < Max.Iter)) { 
    Iter <- Iter + 1 
    x1 <- Newton(f, x0, Delta) 
    TOL <- abs(f(x1)) 
    x0 <- x1 
  } 
  x0 
} 
```

### Finding Roots of a Function

```{r}
Root <- Solve(Square, 1.5, 0.0001, 10^(-6), 10) 
c(Root, Root^2) 
```

### Applying Newton-Raphson to the Score Function of Q

```{r}
Score <- function(Q) 3/Q - 2/(1 - Q) + 1/(4 - Q) + 3/(4 - 3*Q) 
q.hat <- Solve(Score, 0.5, 0.0001, 10^(-6), 10) 
q.hat 
```

### Alternative Root Finding Using `uniroot()`

The `uniroot()` function in R can also find the root of a function over a specified interval:

```{r}
x.lower <- 0.001; x.upper <- 0.999 
uniroot(Score, c(x.lower, x.upper)) 
```

### Numerical Derivative of the Log-Likelihood Function

This works with a numerical derivative of the log likelihood function:

```{r}
Log.L <- function(Q) 3*log(Q) + 2*log(1-Q) - log(4-Q) - log(4-3*Q) 
Score.Num <- function(x) (Log.L(x + Delta/2) - Log.L(x - Delta/2))/Delta 
Delta <- 0.0001 
uniroot(Score.Num, c(x.lower, x.upper)) 
```

### Checking the Score Function at the Estimated Root

```{r}
Score(q.hat) 
```

```{r}
Score(uniroot(Score.Num, c(x.lower, x.upper))$root) 
```

-   How does the Newton-Raphson method compare to `uniroot()`?
-   How does step size (Delta) affect convergence?
-   What are the advantages of using Newton-Raphson in maximum likelihood estimation?
