---
title: "Class 2: Kaplan-Meier"
author: "Yubo Rasmussen"
format: html
editor: visual
---

# Introduction

This notebook demonstrates how to compute and visualize Kaplan-Meier estimates of the survival function using the `survival` package in R. Each example explores a different dataset and scenario to highlight the flexibility of Kaplan-Meier survival analysis.

## Example 1: Time to Discontinuation of a Medical Device

The data represent an experiment studying the time to discontinuation of a medical device. Observations include both complete and censored data, indicated by `Censor`.

### Load the Data

```{r}
library(survival)

# Time to discontinuation (in weeks)
Time <- c(10,13,18,19,23,30,36,38,54,56,59,75,93,97,104,107,107,107)

# Censoring indicator (1 = event occurred, 0 = censored)
Censor <- c(1,0,0,1,0,1,1,0,0,0,1,1,1,1,0,1,0,0)

# Create a survival object
Surv(Time, Censor)
```

**Questions and Answers**:

1\. **What is the purpose of the `Surv()` function?**

2\. **Why do we need the censoring indicator?** \### Kaplan-Meier Estimate

```{r}
# Fit the Kaplan-Meier survival curve
KM.Surv <- survfit(Surv(Time, Censor) ~ 1, conf.type = "plain")

# Summary of the survival object
summary(KM.Surv)
```

**Questions and Answers**:

1\. **What does `~ 1` mean in this context?**

2\. **What does `conf.type = "plain"` mean in this context?**

3\. **What does the Kaplan-Meier estimate represent?**

### Plot the Survival Curve

```{r}
plot(KM.Surv, main = "Time to Discontinuation of Medical Device",
     xlab = "Time (in weeks)", ylab = "Survival Function with 95% CI")
```

**Questions and Answers**:

1\. **Why is it important to visualize the Kaplan-Meier curve?**

2\. **What do the confidence intervals indicate on the plot?**

------------------------------------------------------------------------

## Example 2: Acute Myelogenous Leukaemia (AML) Data

This example uses AML data where: - `1` indicates a relapse occurred. - `0` indicates the patient left the study or was still in remission.

### Load and Explore the Data

```{r}
# Load the aml dataset
library(survival)
aml
attach(aml)

# Create a survival object
Surv(time, status)
```

**Questions and Answers**:

1\. **What does `status` represent in the AML dataset?**

2\. **How does the Kaplan-Meier estimate handle censored patients in this dataset?**

### Kaplan-Meier Estimate by Group

```{r}
# Fit the Kaplan-Meier survival curve by group (x)
fit <- survfit(Surv(time, status) ~ x, data = aml)

# Summary of the survival object
summary(fit)
```

**Questions and Answers**:

1\. **What does stratifying by `x` achieve in this analysis?**

2\. **What key differences might you expect between groups in this example?**

### Plot the Survival Curve

```{r}

plot(fit, lty = 2:3, main = "Kaplan-Meier Survival Curves for AML",
     xlab = "Time (in days)", ylab = "Survival Probability")

# Add legend
legend(100, 0.8, c("Maintained", "Nonmaintained"), lty = 2:3)
```

**Questions and Answers**:

1\. **What does the legend add to this plot?**

2\. **Why are different line types used for groups?**

------------------------------------------------------------------------

## Example 3: Rat Drug Trial

This example involves survival times of rats in a drug trial. Observations include both events and censored data.

### Load the Data

```{r}
# Time to event (in days)
Time <- c(3,4,6,11,11,17,21,24,25,26,30,30,30,30,30)

# Censoring indicator (1 = event occurred, 0 = censored)
Censor <- c(1,1,0,1,1,0,1,0,0,1,0,0,0,0,0)

# Ensure lengths match
length(Time) == length(Censor)

# Create a survival object
Surv(Time, Censor)
```

**Questions and Answers**:

1\. **Why is it important to check that the lengths of `Time` and `Censor` match?**

2\. **What assumptions are made about censored data in this example?**

### Kaplan-Meier Estimate

```{r}
# Fit the Kaplan-Meier survival curve
KM.Surv <- survfit(Surv(Time, Censor) ~ 1, conf.type = "plain")

# Summary of the survival object
summary(KM.Surv)
```

### Plot the Survival Curve

```{r}
plot(KM.Surv, main = "Rat Drug Trial",
     xlab = "Time (in days)", ylab = "Survival Function with 95% CI")
```

**Questions and Answers**:

1\. **What features of the survival curve might indicate the effect of the drug?**
